---
title: "Eigshow"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(MASS)
library(plotly)
```

## Eigshow, Week 1

One of my all-time favourite demonstrations is the `eigshow` concept from MATLAB.
A green vector **x** sits on the unit circle; a blue vector **Ax** shows where the
linear map sends it.  Dragging **x** around traces an ellipse.  Below each matrix
is an interactive animation: press **Play** (or drag the slider) to sweep **x**
around the circle and watch **Ax** trace the ellipse.  The animation pauses and
annotates automatically whenever **Ax** becomes parallel to **x** — that is, when
an eigenvector is found.

---

```{r eigshow-anim, echo=TRUE}
# Animate x sweeping the unit circle; Ax traces the ellipse.
# Pauses at eigenvector angles and labels the eigenvalue.
eigshow_anim <- function(A, title_str = "", n_steps = 90, pause_n = 20) {

  ev      <- eigen(A)
  real_ev <- all(Im(ev$values) == 0)

  theta <- seq(0, 2 * pi, length.out = n_steps + 1)[seq_len(n_steps)]
  dt    <- 2 * pi / n_steps

  # Catalogue each eigenvector direction (both ±v) with a display label
  eig_cat <- if (real_ev) {
    do.call(c, lapply(seq_len(ncol(ev$vectors)), function(j) {
      v   <- Re(ev$vectors[, j])
      lam <- Re(ev$values[j])
      lbl <- paste0("λ", j, " = ", sprintf("%.4g", lam),
                    "  →  eigenvector!")
      lapply(c(1, -1), function(s) {
        list(angle = atan2(s * v[2], s * v[1]) %% (2 * pi), label = lbl)
      })
    }))
  } else list()

  nearest_eig <- function(th) {
    for (ec in eig_cat) {
      d <- min(abs(th - ec$angle), 2 * pi - abs(th - ec$angle))
      if (d < 1.5 * dt) return(ec)
    }
    NULL
  }

  step_eig   <- lapply(theta, nearest_eig)
  step_label <- vapply(step_eig, function(e) if (is.null(e)) "" else e$label,
                       character(1))
  is_eig     <- !vapply(step_eig, is.null, logical(1))

  # Expand: repeat eigenvector frames to create a visual pause
  step_seq <- rep(seq_len(n_steps), times = ifelse(is_eig, pause_n, 1L))
  n_frames <- length(step_seq)

  x_mat  <- rbind(cos(theta), sin(theta))
  Ax_mat <- A %*% x_mat
  lim    <- max(abs(c(x_mat, Ax_mat))) * 1.45

  # Dashed reference lines along each eigenvector (drawn once, always visible)
  eig_shapes <- if (real_ev) {
    lapply(seq_len(ncol(ev$vectors)), function(j) {
      v <- Re(ev$vectors[, j])
      list(type = "line",
           x0 = -lim * v[1], y0 = -lim * v[2],
           x1 =  lim * v[1], y1 =  lim * v[2],
           line  = list(color = "rgba(100,100,100,0.40)",
                        width = 1.3, dash = "dash"),
           layer = "below")
    })
  } else list()

  # One plotly frame per entry in step_seq
  all_frames <- lapply(seq_len(n_frames), function(f) {
    i    <- step_seq[f]
    is_h <- is_eig[i]
    ix   <- seq_len(i)

    cx  <- x_mat[1, i];  cy  <- x_mat[2, i]
    cAx <- Ax_mat[1, i]; cAy <- Ax_mat[2, i]

    # Thicker, darker vectors at eigenvector frames
    vw <- if (is_h) 4   else 2
    xc <- if (is_h) "#005a32" else "#33a02c"
    Ac <- if (is_h) "#08306b" else "#1f78b4"

    annot <- if (is_h) {
      list(list(
        x = -lim * 0.93, y = lim * 0.90, xanchor = "left",
        showarrow = FALSE,
        text = paste0("<b>", step_label[i], "</b>"),
        font = list(color = "#cb181d", size = 15)
      ))
    } else list()

    list(
      name   = as.character(f),
      data   = list(
        list(x = x_mat[1, ix],  y = x_mat[2, ix]),        # trace 0: x trail
        list(x = Ax_mat[1, ix], y = Ax_mat[2, ix]),       # trace 1: Ax trail
        list(x = c(0, cx),  y = c(0, cy),                 # trace 2: x vector
             line   = list(color = xc, width = vw),
             marker = list(size = c(2, 12), color = xc)),
        list(x = c(0, cAx), y = c(0, cAy),                # trace 3: Ax vector
             line   = list(color = Ac, width = vw),
             marker = list(size = c(2, 12), color = Ac))
      ),
      layout = list(annotations = annot)
    )
  })

  # Base figure — four empty traces whose structure matches frame updates
  fig <- plot_ly() |>
    add_trace(x = numeric(0), y = numeric(0),
              type = "scatter", mode = "lines", name = "x trail",
              line = list(color = "#33a02c", width = 1.2)) |>
    add_trace(x = numeric(0), y = numeric(0),
              type = "scatter", mode = "lines", name = "Ax trail",
              line = list(color = "#1f78b4", width = 1.2)) |>
    add_trace(x = numeric(0), y = numeric(0),
              type = "scatter", mode = "lines+markers", name = "x",
              line   = list(color = "#33a02c", width = 2),
              marker = list(size = 10, color = "#33a02c")) |>
    add_trace(x = numeric(0), y = numeric(0),
              type = "scatter", mode = "lines+markers", name = "Ax",
              line   = list(color = "#1f78b4", width = 2),
              marker = list(size = 10, color = "#1f78b4"))

  fig$x$frames <- all_frames

  fig |>
    layout(
      title  = list(text = title_str, x = 0.04, font = list(size = 14)),
      xaxis  = list(range = c(-lim, lim), zeroline = TRUE, showgrid = FALSE,
                    scaleanchor = "y", scaleratio = 1, title = ""),
      yaxis  = list(range = c(-lim, lim), zeroline = TRUE, showgrid = FALSE,
                    title = ""),
      shapes = eig_shapes,
      legend = list(orientation = "h", y = -0.10),
      margin = list(t = 50, b = 100),
      updatemenus = list(list(
        type = "buttons", direction = "left",
        x = 0.04, y = -0.08, xanchor = "left",
        buttons = list(
          list(label  = "&#9654;  Play",
               method = "animate",
               args   = list(NULL, list(
                 fromcurrent = TRUE,
                 frame       = list(duration = 60, redraw = FALSE),
                 transition  = list(duration = 0)
               ))),
          list(label  = "&#9646;&#9646;  Pause",
               method = "animate",
               args   = list(list(), list(
                 mode       = "immediate",
                 frame      = list(duration = 0, redraw = FALSE),
                 transition = list(duration = 0)
               )))
        )
      ))
    ) |>
    animation_slider(
      currentvalue = list(prefix = "Step ", font = list(size = 11)),
      pad          = list(t = 55)
    )
}
```

---

## The First Matrix

When you start `eigshow` in MATLAB, a 2-by-2 matrix appears in the title bar.
There is a green vector **x** and a blue vector **Ax**, the image of **x** under
the map induced by **A**.  The default values are:

```{r matrix-1}
A1 <- rbind(c(1, 3), c(4, 2)) / 4
x  <- c(1, 0)

fractions(A1)
x
fractions(A1 %*% x)
```

### Animation

Press **Play**.  The green dot traces the unit circle; the blue dot traces the
ellipse.  The dashed grey lines show the two eigenvector directions from the
start.  The animation pauses and turns red whenever **Ax** aligns with **x**.

```{r anim-1, out.width="100%"}
eigshow_anim(A1, title_str = "A = [1 3; 4 2] / 4")
```

### Ellipse

As **x** sweeps the unit circle, **Ax** traces an ellipse.  This matrix is *not*
symmetric, so the eigenvectors are **not** the major/minor axes of the ellipse.
A symmetric matrix would align them.

### Eigenvectors

One eigenvector sits in the first/third quadrant (≈ 60° from horizontal); the
blue vector is *longer* than the green, so the eigenvalue is greater than 1.
The other is found when **Ax** points *opposite* to **x** (negative eigenvalue),
in the second/fourth quadrant at ≈ 45°.

```{r eigen-1}
fractions(eigen(A1)$values)
fractions(eigen(A1)$vectors)
```

---

## A Second Matrix

Swap the top two elements of the first matrix.

```{r matrix-2}
A2 <- rbind(c(3, 1), c(4, 2)) / 4
fractions(A2)
```

```{r anim-2, out.width="100%"}
eigshow_anim(A2, title_str = "A = [3 1; 4 2] / 4")
```

The ellipse is noticeably more elongated.  Both eigenvalues are positive, so
**Ax** goes around the ellipse in the *same* direction as **x** around the
circle.  (The first matrix had one negative eigenvalue; **x** and **Ax** then
travel in opposite senses.)

```{r eigen-2}
eigen(A2)$values
```

---

## A Third Matrix

Swap the bottom two elements of the first matrix and flip one sign.

```{r matrix-3}
A3 <- rbind(c(3, 1), c(-2, 4)) / 4
fractions(A3)
```

```{r anim-3, out.width="100%"}
eigshow_anim(A3, title_str = "A = [3 1; -2 4] / 4")
```

No pause, no annotation, no dashed lines — the animation just keeps going.
Trying to make **Ax** parallel to **x** is impossible on the real plane.
This matrix has no real eigenvalues or eigenvectors; they are complex.

```{r eigen-3}
eigen(A3)$values   # complex conjugate pair
```

---

## More Later

Further examples will appear in the next set of notes.

---

*Adapted from Cleve Moler, © 2013 The MathWorks, Inc.*
